\LARGE \bf Explicit Knowledge and the Deliberative Layer: Lessons Learned

### 2013-07-08 paper.tex
IROS review: style + grammar

We have chosen to tackle the challenge from several ends: human-aware
navigation and motion planning (\cite{Mainprice2011}), situation assessment
coupled with motion planning (\cite{Mainprice2012}), projection of
``mightabilities'' that anticipates what surrounding agents may do
(\cite{Pandey2011}), and the design and deployment of a pervasive
\emph{knowledge-centric} software architecture.

The last item is the focus of this paper: what does ``knowledge-centric''
architecture means, and what are its benefits and drawbacks.

@@

In the next section, we first introduce the deliberative layer of our robots at
the functional level. It represents what we call a \emph{knowledge-oriented}
architecture, built around a central knowledge management tool called the {\sc
Oro} server.

The following sections go over each of the facets of this deliberative layer:
how knowledge is produced asynchronously from geometric reasoning, how the
robot conducts grounded multi-modal interaction with humans, how knowledge is
used by decision-making components like the robot controller or the task
planner, and finally, how our knowledge-centric architecture enables the
implementation of specific internal cognitive processes.

@@

This architecture moves away from standard layered approaches. Interactions
between components are mostly bidirectional and, from the software components
point of view, we do not introduce layers of abstraction (we do, however, have
access to the lower level modules of the robot to execute actions, but all
cognition-related modules reside at the same level). This is especially visible
for the dialogue input processing. This component does not simply act as an
alternative perceptual input to the symbolic database, but also actively
queries previously acquired knowledge to disambiguate and validate the newly
created symbolic knowledge (see section~\ref{sect|com}).

@@

This interaction between knowledge and actions is also central to our approach
(as for any cognitive system), but task representation and task execution is
not seen as a monolithic, central function: it is one of the activities of the
robot, actually split between communication components (that can acquire
desires from interaction with agents, amongst other things) and an execution
controller that may decide to take an incoming desire into account to create
its own internal goals. The controller generates and controls intentions from
these goals with the help of a symbolic task planner, that has also direct
access to the knowledge base.

@@

In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
\emph{semantic blackboard}: one central server (the {\sc Oro}
server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
deliberative components. It conversely exposes an API to query the knowledge
base.

@@

This knowledge model also largely exclude representation of continuous
phenomena (like time) or uncertain phenomena. When required, these have to be
managed by dedicated components.

Amongst the features of {\sc Oro}, the system can dynamically creates
new independent knowledge models (\ie independent instantiations of the
ontology). This feature is used for instance to store separately the beliefs of
each of the agents the robot interacts with (see section~\ref{sect|tom}).

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module (as named on Fig.~\ref{fig|archi}) called SPARK (\emph{SPAtial
Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a situation assessment
reasoner that generates symbolic knowledge from the geometry of the
environment with respect to relations between objects, robots and humans,
also taking into account the different perspective that each agent has on
the environment (Fig.~\ref{fig|spark-oro}).

SPARK is an \emph{amodal} geometric model of the environment that serves both
as basis for the fusion of the perception modalities and as bridge with the
symbolic layer. This geometric model is continuously updated at run-time by the
robot based on its sensors. In our experiments, objects are identified and
localised through 2D barcodes.  On the other hand, the human pose is tracked
with a Kinect-like device (assisted by motion capture to accurately track the
head motion, which is required to compute what the human is looking at).

@@

Absolute spatial relations (\concept{isOn}, \concept{isIn}, \concept{isNextTo},
\concept{isAbove}...) are computed for every pairs of object, taking into
account relative sizes for relations like \concept{isNextTo}. Other
agent-independent properties like \concept{isMoving} are also computed.

@@

SPARK also computes affordances: \concept{sees}, \concept{looksAt},
\concept{pointsAt}, \concept{canReach}. Visibility (with two levels
corresponding to the field of view and the field of attention) and pointing are
computed by placing virtual cameras in the SPARK 3D model, while reachability
relies on inverse kinematics. Several filters (based on hystersis, expected
object behaviour, etc.) allow to partially account for perception
noise~\cite{Warnier2012a}.

@@

\subsection{Natural language grounding}

Natural language processing is one of the fields of human-robot interaction
where the introduction of the semantic layer has been most beneficial.
We detail in~\cite{Lemaignan2011a} the techniques and the tool called
\texttt{dialogs} that we have developed for English language parsing and
grounding, along with verbalisation and (minimalist) dialogue management.

@@

The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal}, interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
tired}).

\subsection{Limits of disambiguation at semantic level}

@@

Note that, while the system benefits from the complementary modalities, they
are not all required. The dialogue system would for instance run equally well
with only the verbal modality.

@@

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. In its simplest
form, a module can register an event by passing through an event pattern, which
is basically a partial statement like \stmt{? type Elephant}, and a callback.
Every time a new instance of elephant appears in the knowledge base, the
callback is triggered.

@@

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), it can
focuses on the semantic of the desire.

Note also that we take advantage of the reasoning capabilities of the system:
for example, the goal of the action (\stmt{action hasGoal myself}) may well not
have been explicitly asserted, but inferred by the reasoner based on other
assertions.

@@

This simplistic measurement needs to be refined to take into account the fact
that the raw number of RPC calls to the knowledge base only imperfectly
reflects the actual cognitive processes running on, but it opens interesting
perspective nevertheless. It would be interesting for example to compare
and look for correlations with the cognitive load of a human partner (measured
via EEG of qualitative feedback).

@@

The {\sc Oro} server also feature a mechanism to mimic simple forms of
biological memory.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

@@

We rely however on this short term memory for a particular use-case:
\emph{active concept}. Some modules, like the natural language processor, use
the {\tt short term} memory profile to mark for a few seconds important
concepts that are currently manipulated by the robot. For example, if a human
asks the robot: ``Give me all red objects'', the human, the \concept{Give}
action, and each red objects that are found are successively marked as
\emph{active concepts} by inserting statements like \stmt{human type
ActiveConcept} in the short-term memory (which can be considered, in this case,
to be a working memory). We use this feature to give a (visual) feedback to the
users.

@@

We conclude this article with three different perspectives on the knowledge-centric
techniques we have presented here: first, the \emph{architect perspective}:
what are the pro and cons of our approach from an engineer point of view. Then,
the \emph{logician perspective}: does the promise of automated reasoning for
robots really comes to fruition with ontologies? And finally the
\emph{cognitician perspective} or how explicit knowledge manipulation endows
our robots with new cognitive capabilities.

@@

Since these statements are anchored in the robot's pool of knowledge, they
actually convey shared and unambiguous \emph{meaning} through the system. This
has one important consequence from the perspective of the system architecture:
modules interfaces are only defined in term of knowledge consumption and/or
production. This brings good decoupling properties. For instance, the output of
the dialog module is a new \emph{situation desired} by the speaker. This module
can be transparently replaced or completed by any other module that produces
the same chunks of knowledge. Only a thin protocol layer remains to get the
modules to interact together. This allows us to implement (otherwise difficult)
merging of radically different perception modalities like vision (the robot
sees an object on a table: the output would look like \stmt{obj1 isOn table1})
and verbal description (a human says to the robot the object on the table
is red: it leads to a query \stmt{? isOn table1} and then a new assertion:
\stmt{obj1 hasColor red}).

Another consequence is that the communication channels are {\it
de facto} defined by the semantics of the information they
convey. This allows the system designers to think in term of which module
produces or needs which knowledge, instead of which module produces which
service. We found this approach to be very helpful when implementing cognitive
functions like the theory of mind, the perspective taking or the natural
language grounding.

Lastly, another good property of knowledge-based communication between module
is the \emph{cognitive observability} of the system: by logging queries between
modules (which is even easier in our case since we rely on a central knowledge
management), one can trace the interactions between the robot systems at a high
abstraction level. We will come back below on the significance of this property
from the cognitive point of view.

The knowledge-centric architecture also has downsides. We can name here a few.

@@

On the sensing side also we have in certain cases to bypass the semantic layer
(extension of the symbolic task planner with geometric constraints is one
example).

And obviously, many aspects like uncertainty management, modal logics,
non-monotonic reasoning,... are not well addressed with the Description Logic
formalism, and remain to be explored to gain the level of expressiveness
required to cover many of the more complex human-robot interaction scenarii.

@@

This example of reasoning, quoted from Moritz Tenorth, is a good example of
simple yet non-trivial reasoning. As a matter of fact, only very few of such
prototypical reasoning cases where positively identified in our scenarii and experiments
(and consequently implemented as rules in {\sc Oro}).

@@

The experiments we have conducted are also likely to have too simplistic
semantics to let complex reasoning needs to emerge. Scenarii with more complex
semantics would be desirable to better stress the expressiveness and inference
abilities provided by Description Logics.

Is reasoning at the knowledge level immature or even superfluous, then? Not so:
hundreds of trivial (from a human point of view) inferences are continuously
produced by the system (translating inheritance relations, domain/range
constraints, transitivity, etc.) and encode a large amount of common-sense
knowledge that can not be otherwise asserted. These trivial inferences are all
the more important that an expressive knowledge representation language is
used: when a language like OWL allows to directly represent high-level concepts
like partitions, cardinality restrictions, properties' ranges and domains, it
leads to a more implicit (because more abstract) description of the vocabulary
that in turn requires more underlying reasoning. With the progress in the
understanding of the relations between expressiveness and (tractable)
satisfiability, along with the progress of reasoners, more and more of the
inferences do not need to be explicit anymore, and consequently are delegated
to the reasoner.

>>>>

We have chosen to tackle the challenge from several ends: human-aware
navigation and motion planning~\cite{Mainprice2011}, situation assessment
coupled with motion planning~\cite{Mainprice2012}, projection of
``mightabilities'' that anticipates what surrounding agents may
do~\cite{Pandey2011}, and the design and deployment of a pervasive
\emph{knowledge-oriented} software architecture.

The last item is the focus of this paper: what ``knowledge-oriented''
architecture means, and what are its benefits and drawbacks.

@@

In the next section, we first introduce the deliberative layer of our robots at
the functional level. It represents what we call a \emph{knowledge-oriented}
architecture, built around a central knowledge management tool called the
OpenRobot Ontology ({\sc Oro}) server.

The following sections go over each of the facets of this deliberative layer:
how knowledge is produced asynchronously from geometric reasoning, how the
robot conducts grounded multi-modal interaction with humans, how knowledge is
used by decision-making components like the robot controller or the task
planner, and finally, how our knowledge-oriented architecture enables the
implementation of specific internal cognitive processes.

@@

This architecture moves away from standard layered approaches. Interactions
between components at the deliberative level are mostly bidirectional and we do
not introduce layers of abstraction amongst software components\footnote{We do
have lower-level modules to execute actions or manage sensors, but all
cognition-related modules reside at the same level.}. This is nicely
illustrated by the dialogue input processing. This component does not simply
act as an alternative perceptual input to the symbolic database: it also
actively queries previously acquired knowledge to disambiguate and validate the
newly created symbolic knowledge (see section~\ref{sect|com}).

@@

This interaction between knowledge and actions is also central to our approach
(as for any cognitive system), but task representation and task execution is
not seen as a monolithic, central function: it is one of the activities of the
robot, actually split between communication components (that can acquire
desires from interaction with agents, amongst other things) and an execution
controller that may decide to take an incoming desire into account to create
its own internal goals. The controller generates and manages intentions from
these goals with the help of a symbolic task planner, that also has direct
access to the knowledge base.

@@

In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
\emph{semantic blackboard}: one central server (the {\sc Oro}
server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
deliberative components. It conversely exposes a {\tt json}-based RPC API to
query the knowledge base~\cite{lemaignan2012kbapi}.

@@

This knowledge model also largely exclude representation of continuous
phenomena (like time) or uncertain phenomena. When required, these are managed
inside the corresponding components, and are not exposed at the system level.

Amongst the features of {\sc Oro}, the system can dynamically create
new independent knowledge models (\ie independent instantiations of the
ontology). This feature is used for instance to store separately the beliefs of
each of the agents the robot interacts with (see section~\ref{sect|tom}).

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called SPARK (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Fig.~\ref{fig|spark-oro}),
also taking into account the different perspective that each agent has on the
environment.

SPARK is an \emph{amodal} geometric model of the environment that serves both
as basis for the fusion of the perception modalities and as bridge with the
symbolic layer. This geometric model is continuously updated at run-time by the
robot based on its sensors (in our experiments, objects are identified and
localised through 2D barcodes, while humans are tracked with Kinect-like
devices, optionally assisted by motion capture to accurately track the head
motion, which is required to compute what the human is looking at).

@@

Absolute spatial relations (\concept{isOn}, \concept{isIn}, \concept{isNextTo},
\concept{isAbove}...) are computed for every pair of objects, taking into
account relative sizes for relations like \concept{isNextTo}. Other
agent-independent properties like \concept{isMoving} are also computed.

@@

SPARK also computes affordances: \concept{sees}, \concept{looksAt},
\concept{pointsAt}, \concept{canReach}. Visibility (with two levels
corresponding to the field of view and the field of attention) and pointing are
computed by placing virtual cameras in the SPARK 3D model, while reachability
relies on inverse kinematics. Several filters (based on hysteresis, expected
object behaviour, etc.) allow to partially account for perception
noise~\cite{Warnier2012a}.

@@

\paragraph*{Natural language grounding}

Natural language processing is one of the fields of human-robot interaction
where the introduction of the semantic layer has been most beneficial.
We detail in~\cite{Lemaignan2011a} the techniques and the tool called
\texttt{dialogs} that we have developed for natural English language parsing and
grounding, along with verbalisation and (minimalist) dialogue management.

@@

The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal}), interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
tired}).

\paragraph*{Limits of disambiguation at semantic level}

@@

Note that, while the system benefits from the complementary modalities, they
are not all required. The dialogue system would for instance happily run with
only the verbal modality, at the cost of weaker interaction.

@@

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. In its simplest
form, a module can subscribe to an event by passing through an event pattern
(in its basic form, a partial statement like \stmt{? type Elephant}) and a
callback.  Every time a new instance of elephant appears in the knowledge base,
the callback is triggered.

@@

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), it can
focus on the semantic of the desire.

Note also that we take advantage of the reasoning capabilities of the system:
for example, the goal of the action (\stmt{action hasGoal myself}) may not be
explicitly asserted, but inferred by the reasoner based on other assertions.

@@

This simplistic measurement needs to be refined to take into account the fact
that the raw number of RPC calls to the knowledge base only imperfectly
reflects the actual cognitive processes running on, but it opens interesting
perspective nevertheless. It would be interesting for example to compare
and look for correlations with the cognitive load of a human partner (measured
via EEG or qualitative feedback).

@@

The {\sc Oro} server also feature a mechanism to mimic minimalistic forms of
biological memory.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

@@

We rely however on this short term memory for a particular use-case:
\emph{active concepts}. Some modules, like the natural language processor, use
the {\tt short term} memory profile to mark for a few seconds important
concepts that are currently manipulated by the robot. For example, if a human
asks the robot: ``Give me all red objects'', the human, the \concept{Give}
action, and each red objects that are found are successively marked as
\emph{active concepts} by inserting statements like \stmt{human type
ActiveConcept} in the short-term memory (which can be considered, in this case,
to be a working memory). We use this feature to trace certain knowledge-related
processes.

@@

We conclude this article with three different perspectives on the knowledge-oriented
techniques we have presented here: first, the \emph{architect perspective}:
what are the pro and cons of our approach from an engineer's point of view. Then,
the \emph{logician perspective}: does the promise of automated reasoning for
robots really comes to fruition with ontologies? And finally the
\emph{cognitician perspective} or how explicit knowledge manipulation endows
our robots with new cognitive capabilities.

@@

Since these statements are anchored in the robot's pool of knowledge, they
actually convey shared and unambiguous \emph{meaning} through the system. This
has one important consequence from the perspective of the system architecture:
interfaces between modules are only defined in term of knowledge consumption and/or
production. This brings good decoupling properties. For instance, the output of
the dialog module is a new \emph{situation desired} by the speaker. This module
can be transparently replaced or completed by any other module that produces
the same chunks of knowledge. Only a thin protocol layer remains to get the
modules to interact together. This allows us to implement (otherwise difficult)
merging of radically different perception modalities like vision (the robot
sees an object on a table: the output would look like \stmt{obj1 isOn table1})
and verbal description (a human says to the robot the object on the table
is red: it leads to a query \stmt{? isOn table1} and then a new assertion:
\stmt{obj1 hasColor red}).

Another consequence is that the communication channels are {\it
de facto} defined by the semantics of the information they
convey. This allows the system designers to think in term of which module
produces or needs which knowledge, instead of which module produces which
service. We found this approach to be helpful when implementing cognitive
functions like the theory of mind, the perspective taking or the natural
language grounding.

Lastly, another good property of knowledge-based communication between module
is the \emph{cognitive observability} of the system: by logging queries between
modules (which is even easier in our case since we rely on a central knowledge
management), one can trace the interactions between the robot subsystems at a high
abstraction level. We will come back below on the significance of this property
from the cognitive point of view.

The knowledge-oriented architecture also has downsides. We can name here a few.

@@

Also on the sensing side, we have in certain cases to bypass the semantic layer
(extension of the symbolic task planner with geometric constraints is one
example).

Many aspects like uncertainty management, modal logics,
non-monotonic reasoning are not well addressed with the Description Logic
formalism, and remain to be explored to gain the level of expressiveness
required to cover many of the more complex human-robot interaction scenarios.

@@

This example of reasoning, quoted from Moritz Tenorth, is a good example of
simple yet non-trivial reasoning. As a matter of fact, only very few of such
prototypical reasoning cases were positively identified in our scenarios and experiments
(and consequently implemented as rules in {\sc Oro}).

@@

The experiments we have conducted are also likely to have too simplistic
semantics to let complex reasoning needs to emerge. Scenarios with more complex
semantics would be desirable to better stress the expressiveness and inference
abilities provided by Description Logics.

Is reasoning at the knowledge level immature or even superfluous, then? Not so:
hundreds of trivial (from a human point of view) inferences are continuously
produced by the system (translating inheritance relations, domain/range
constraints, transitivity, etc.) and encode a large amount of common-sense
knowledge that can not be otherwise conveniently asserted. These trivial
inferences are all the more important that an expressive knowledge
representation language is used. Since a language like OWL allows it to
directly represent high-level concepts like partitions, cardinality
restrictions, properties' ranges and domains, it leads to a more
\emph{implicit} description of the available knowledge.  This in turn pushes
for a lot of ``trivial'' reasoning to make the knowledge pool explicit, and
hence largely reusable by the deliberative subsystems. With the progress in the
understanding of the relations between expressiveness and (tractable)
satisfiability, along with the progress of reasoners, more and more of the
assertions do not need to be explicit anymore, and consequently are delegated
to the reasoner.
